Demes: a specification for interchanging complex demography

### 2021-08-06 paper.tex

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}

@@

\abstract{
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being estimated.
Demographic models of current interest (consisting of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs) can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambigous
graphical model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
}

@@

Simulation is a core tool in population genetics, and a
large number of different methods have been developed
over the past three decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation method~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and (potentially) also allow for documentation in the
form of comments. However, there is no standard format shared across
simulators, and the details of the demographic models are not separated
from other simulation parameters. The APIs used by individual simulators
to define demographic models are also not appropriate as an interchange
standard because of the dependency on both the programming language involved,
and the specific details of the simulator in question.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissable ranges. This schema, and its
accompanying documentation, fully describe the entities in the model and their
relationships and the required behaviour of implementations. Since the
schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

@@

In Demes, demographic models consist of one or more interacting populations (or
``demes''; to avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, in the understanding that the
terms are interchangeable for our purposes). Populations are defined as a
collection of exchangeable individuals that exist for a specified period of
time, following a well-defined set of rules regarding population sizes, mating
systems, and so on. Ancestor/descendant relationships between populations can
be defined, as well as continuous or instantaneous migration between coexisting
populations.

@@

Within a population, individuals are assumed to be exchangeable, but there are
also parameters for nonrandom mating, such as selfing or cloning rates, which
give the probability that offspring are generated from one generation to the
next by self-fertilization or cloning of an individual. Cloning and selfing
rates can take any value between zero and one, and if both are specified then
the selfing rate is the probability an offspring is generated through
self-fertilization given it was produced sexually (i.e. with probability one
minus the cloning rate). We omit any details regarding selection, as those would
fall outside the scope of demographic model specification, but...
\aprcomment{If selection is placed on top of the demographic model, this implies
that we are living in a soft selection world. Though sizes could just as easily
be interpreted as carrying capacities in a hard selection model.}

@@

The high-level description of Demes models is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (see Figure~\ref{fig-example} for examples). YAML is a data
serialisation language with an emphasis on simplicity, and interoperates well
with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalance with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig-example} shows an example Demes model expressed in both the
high and low-level forms (both in terms of YAML syntax). For values that repeat
across fields, defaults may be used to implicity assign default values to
fields of the given type.  A default is superceded by an explicitly provided
value if given. Other implicit values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided, it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size.

@@

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem). We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

@@

\subsection*{Scope of the specification}
\jkcomment{This section is very rough.}

@@

The Demes specification is by design simple and static---we wish to
unambiguously describe a demographic model with fully specified
and fixed parameters. This simplicity means that we cannot directly
specify distributions or estimated confidence intervals
for those parameters. While it is not difficult to imagine extending
the specification in ways that would allows us to do this, it would
greatly increase the complexity of parsers and would likely be of
little practical use in any case. This is because the parameters of
demographic models are typically tightly coupled, and cases in which
distributions for different parameters can be simply described are rare.
In this situation, the simplest way to describe estimated
distribution is simply to list a large number of samples from
the posterior. While writing out a large number of Demes models in
YAML format may seem inefficient, it can in fact be a compact
way to describe these distributions.
For example, consider a one-population model with piecewise-constant sizes over
20 epochs which has $\sim$~40 free parameters---the \texttt{start\_size} and
\texttt{end\_time} values for each epoch. Supposing we sample 50,000 models
from the posterior distribution, the resulting multi-document YAML file is
45~MiB.
This format compresses down to 8.4~MiB when gzipped or 6.2~MiB
when compressed with LZMA2, which is on par with an equivalent binary
representation of the free parameters
($40\times50000\times4~\text{bytes} \approx 7.6~\text{MiB}$).

Similarly, one might be interested in running simulations in which
the demographic model parameters are drawn from a distribution, e.g.,
in ABC inference~\citep{beaumont2002approximate}. Allowing for
the many distributions of interest would greatly increase the complexity of
parsers, with a relatively limited benefit. It would be far simpler
and more powerful to write some code to sample directly from the
distributions of interest, using, for example the Demes Python
API.
\jkcomment{This is garbled, but we get the point? You're probably never
going to be doing ABC from a completely general Demes model as input.
There will be some basic model that you have in mind, that you're generating
distributions of parameters around. So, what's the point in writing
general machinery to do this sampling, when you could write it yourself in
10 lines of Python?}
\jkcomment{Key point: if there was a concise way of writing down the joint
distribution of all the parameters, we wouldn't be doing simulations in the
first place.}

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals. Finally, as Demes assumes exchangeability
of individuals within populations, it not provide a way to define geographic
spatial structure within populations.

@@

% Add a bit more space between rows.
\renewcommand{\arraystretch}{1.5}
\begin{table}
\begin{center}
\begin{tabular}{lp{12cm}}
\toprule
\multicolumn{2}{l}{Software infrastructure}\\
\midrule
\texttt{demes-python} &
    A Python library for loading, saving, and working with
    Demes models. Includes conversion utilities for existing
    demographic model descriptions such as
    \texttt{ms}~\citep{hudson2002generating}.\\

\texttt{demes-c} &
    A C library for parsing Demes YAML descriptions.
    (\url{https://github.com/grahamgower/demes-c.git}). \\

\texttt{demesdraw} &
    A Python library for visualising Demes models
    (\url{https://github.com/grahamgower/demesdraw.git}). \\

@@

The difficulty of describing complex demographic models
for population genetic simulations has long been acknowledged
~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
and a number of different methods have been proposed to
make such simulations more accessible and less error prone.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
ewing2010msms,zhou2018popdemog}. However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the specific syntax
of a given simulator. Another approach that has been taken to
simplify running such simulations is to provide an API in a
scripting language such as Python or R, providing a flexible
interface either to a built-in simulation
engine~\citep{thornton2014cpp,kelleher2016efficient,haller2017flexible}
% jk: are/were there other simulation frontends out there?
or as a frontend for other simulators~\citep{staab2016coala}.
While APIs have many advantages and have grown in popularity in
recent years, the demographic model descriptions are tied
to the details of both the underlying simulation engines
and the programming language in question.

Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been ad-hoc, and conversions between the many different ways
of expressing such models is both labour intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab-software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that either produce or consume this format.

>>>>

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}

@@

\abstract{
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being estimated.
Demographic models of current interest (consisting of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs) can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
graphical model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
}

@@

Simulation is a core tool in population genetics, and a
large number of different methods have been developed
over the past three decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and (potentially) also allow for documentation in the
form of comments. However, there is no standard format shared across
simulators, and the details of the demographic models are not separated
from other simulation parameters. The APIs used by individual simulators
to define demographic models are also not appropriate as an interchange
standard because of the dependency on both the programming language involved,
and the specific details of the simulator in question.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissible ranges. This schema, and its
accompanying documentation, fully describe the entities in the model and their
relationships and the required behaviour of implementations. Since the
schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', by which we mean a grouping of individuals that is convenient
for modelling \citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable for our purposes. Populations are defined as a
collection of exchangeable individuals that exist for a specified period of
time, following a well-defined set of rules regarding population sizes, mating
systems, and so on. Ancestor/descendant relationships between populations can
be defined, as well as continuous or instantaneous migration between coexisting
populations.

@@

Within a population, individuals are assumed to be exchangeable, but there are
also parameters for nonrandom mating, such as selfing or cloning rates, which
give the probability that offspring are generated from one generation to the
next by self-fertilisation or cloning of an individual. Cloning and selfing
rates can take any value between zero and one, and if both are specified then
the selfing rate is the probability an offspring is generated through
self-fertilisation given it was produced sexually (i.e. with probability one
minus the cloning rate). We omit any details regarding selection, as those would
fall outside the scope of demographic model specification, but...
\aprcomment{If selection is placed on top of the demographic model, this implies
that we are living in a soft selection world. Though sizes could just as easily
be interpreted as carrying capacities in a hard selection model.}

@@

The high-level description of Demes models is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (see Figure~\ref{fig-example} for examples). YAML is a data
serialisation language with an emphasis on simplicity, and interoperates well
with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig-example} shows an example Demes model expressed in both the
high and low-level forms (both in terms of YAML syntax). For values that repeat
across fields, defaults may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Other implicit values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided, it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size.

@@

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

@@

\subsection*{Scope of the specification}
\subsubsection*{Static models, not parameterised models}
\jkcomment{This section is very rough.}

@@

The Demes specification is static by design---we wish to
unambiguously describe a demographic model with a concrete set
of parameters. This simplicity means that we cannot directly
specify parameter distributions or estimated confidence intervals
for those parameters. While it is not difficult to imagine extending
the specification in ways that would allow this, it's not immediately
clear that the benefits are worth the increased parser complexity.

\ggcomment{
Possible reasons to advocate for non-static features are as follows.
\begin{enumerate}
\item Specifying prior distributions of parameters for an ABC.
\item Specifying upper and lower bounds of parameters and inter-parameter
      constraints for input to an optimisation-based inference.
\item Specifying joint distribution, or confidence intervals, for the output of
      an inference procedure.
\item Simulating from some arbitrary joint distribution that was inferred
      previously. E.g. using an ABC posterior distribution of models to
      train a neural network.
\item Extending someone else's static model to reinfer parameters,
      or add new parameters.
\end{enumerate}
}

The parameters of
demographic models are typically tightly coupled, and cases in which
distributions for different parameters can be simply described are rare.
In this situation, the simplest way to describe an estimated
distribution is to list a large number of samples from
the posterior. While writing out a large number of Demes models in
YAML format may seem inefficient, it can in fact be a compact
way to describe these distributions.
For example, consider a one-population model with piecewise-constant sizes over
20 epochs which has $\sim$~40 free parameters---the \texttt{start\_size} and
\texttt{end\_time} values for each epoch. Supposing we sample 50,000 models
from the posterior distribution, the resulting multi-document YAML file is
45~MiB.
This format compresses down to 8.4~MiB when gzipped or 6.2~MiB
when compressed with LZMA2, which is on par with an equivalent binary
representation of the free parameters
($40\times50000\times4~\text{bytes} \approx 7.6~\text{MiB}$).

Similarly, one might be interested in running simulations in which
the demographic model parameters are drawn from a distribution, e.g.,
in ABC inference~\citep{beaumont2002approximate}.
Other inference procedures based on optimising a loss function
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
(\ggcomment{more refs? MCMC? neural networks?})
need users to specify parameter bounds,
and possibly non-linear constraints between parameters.
Indeed, the choice of how to parameterise a model could be important for
some inference methods (e.g. absolute times versus relative times between events).

Implementing the many distributions of interest, and supporting a general
way to describe a model's free parameters, would greatly increase the
complexity of parsers, with relatively limited benefit to most users.
It's unlikely that Demes could be made sufficiently
flexible without considerable effort to implement many features of
general-purpose programming languages, such as variables, arithmetic,
and flow control.
Such use cases are therefore better served by writing model-generating
functions in an existing programming language, for example
using the Demes Python API.
To modify an existing static Demes model, perhaps to infer new parameters,
this would unfortunately require reimplementing the model.
However, there exist many templating solutions for YAML and JSON that are
specifically designed for extending static data in arbitrarily complex
(\ggcomment{refs: YTT, Jsonnet, CUE, Dhall}),
offering intriguing possibilities.

\subsubsection*{Population-level features, not genome features}
Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals. Finally, as Demes assumes exchangeability
of individuals within populations, it doesn't provide a way to define geographic
spatial structure within populations.

@@

% Add a bit more space between rows.
\renewcommand{\arraystretch}{1.5}
\begin{table}
\begin{center}
\begin{tabular}{lp{12cm}}
\toprule
\multicolumn{2}{l}{Software infrastructure}\\
\midrule
\texttt{demes-python} &
    A Python library for loading, saving, and working with
    Demes models. Includes conversion utilities for existing
    demographic model descriptions such as
    \texttt{ms}~\citep{hudson2002generating}
    (\url{https://github.com/popsim-consortium/demes-python}).\\

\texttt{demes-c} &
    A C library for parsing Demes YAML descriptions.
    (\url{https://github.com/grahamgower/demes-c}). \\

\texttt{demesdraw} &
    A Python library for visualising Demes models
    (\url{https://github.com/grahamgower/demesdraw}). \\

@@

The difficulty of describing complex demographic models
for population genetic simulations has long been acknowledged
~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
and a number of different methods have been proposed to
make such simulations more accessible and less error prone.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
ewing2010msms,zhou2018popdemog}. However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the specific syntax
of a given simulator. Another approach that has been taken to
simplify running such simulations is to provide an API in a
scripting language such as Python or R, providing a flexible
interface either to a built-in simulation
engine~\citep{thornton2014cpp,kelleher2016efficient,haller2017flexible}
% jk: are/were there other simulation frontends out there?
or as a front end for other simulators~\citep{staab2016coala}.
While APIs have many advantages and have grown in popularity in
recent years, the demographic model descriptions are tied
to the details of both the underlying simulation engines
and the programming language in question.

Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been ad-hoc, and conversions between the many different ways
of expressing such models is both labour intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab-software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that consume, transform, or produce this format.
